Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies

### 2012-07-16 MIMOSA.tex
updated refs
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\section{Introduction}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers). More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell (\textbf{cite CYTOF}). Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize and subset these cells. A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes6. While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability (add references). This heterogeneity may carry important information, thus single cell expression data should not be analyzed in the same fashion as cell-population level data. Special treatment of single cell level data, which preserves information about population heterogeneity, is warranted in general. For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

>>>>

\section{Introduction}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single cell measurement of multiple surface and intracellular markers (up to 17 markers). More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous4,5 and could further characterize and subset these cells. A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes6. While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability (add references). This heterogeneity may carry important information, thus single cell expression data should not be analyzed in the same fashion as cell-population level data. Special treatment of single cell level data, which preserves information about population heterogeneity, is warranted in general. For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\cite{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.
